---
title: "metagenomics case study- HMP"
author: "Kushal K Dey, Joyce HSiao, Joe Paulson"
date: "August 19, 2015"
output: html_document
---

First load the HMP RData and extracting the counts table and the phenotype/ metadata tables.

```{r data_load, echo=TRUE, eval=TRUE, warning=FALSE, message=FALSE}
library(metagenomeSeq)
library(maptpx)
library(tsne)
library(qtlcharts)
load('../../data/HMPvs35_nt100_MRexperiment.rdata')
data <- t(MRcounts(hmp));
metadata <- pData(hmp);
otu_data <- featureData(hmp)
```

Extract the data related to the hands. We first extract those sample indices where the **bodysupersite** label is **Skin** and it seemed they all corresponded to left hand or right hand.

```{r, hands_data, echo=TRUE, eval=TRUE}
hands_data <- data[which(metadata$HMPbodysupersite=="Skin"),];
hands_metadata <- metadata[which(metadata$HMPbodysupersite=="Skin"),];
str(metadata)
```

## Crease vs Fossa (GoM model)

```{r echo=TRUE, eval=FALSE}
hands_gom <- CountClust::FitGoM(hands_data, K=c(2,4,6), tol=0.1);
save(hands_gom, file="../rdas/hands_gom_fit.rda")
```

### K=2

```{r echo=TRUE, eval=TRUE}
hands_gom <- get(load("../rdas/hands_gom_fit.rda"));
omega <- hands_gom$clust_2$omega
theta <- hands_gom$clust_2$theta
```


```{r}
annotation <- data.frame(
    sample_id = paste0("X", 1:nrow(omega)),
    tissue_label = factor(hands_metadata$HMPbodysubsite) ) 


StructureGGplot(omega = omega,
                annotation = annotation,
                palette = RColorBrewer::brewer.pal(8, "Accent"),
                yaxis_label = "hands body subsite",
                order_sample = TRUE,
                axis_tick = list(axis_ticks_length = .1,
                axis_ticks_lwd_y = .1,
                axis_ticks_lwd_x = .1,
                axis_label_size = 10,
                axis_label_face = "bold"))


```

### K=4

```{r echo=TRUE, eval=TRUE}
omega <- hands_gom$clust_4$omega
theta <- hands_gom$clust_4$theta
```


```{r}
annotation <- data.frame(
    sample_id = paste0("X", 1:nrow(omega)),
    tissue_label = factor(hands_metadata$HMPbodysubsite) ) 


StructureGGplot(omega = omega,
                annotation = annotation,
                palette = RColorBrewer::brewer.pal(8, "Accent"),
                yaxis_label = "hands body subsite",
                order_sample = TRUE,
                axis_tick = list(axis_ticks_length = .1,
                axis_ticks_lwd_y = .1,
                axis_ticks_lwd_x = .1,
                axis_label_size = 7,
                axis_label_face = "bold"))


```


The $K=4$ seems to differentiate between crease and fossa, but for $K=2$, the clustering between fossa and crease was not obvious.

We do a PC plot to see if this pattern shows up in the PCs. It does seem  that crease and fossa do cluster sort of together. 

```{r pca, echo=TRUE, eval=TRUE, cache=TRUE}
pr <- prcomp(log(hands_data+0.5)%*%t(log(hands_data+0.5)));
plot(pr)
```

The plot of PC1 vs PC2 is as follows

```{r pca_12, echo=TRUE, eval=TRUE}
suppressWarnings(suppressMessages(iplot(pr$x[,1], pr$x[,2], as.numeric(as.factor(type_bodysite)),type_bodysite)))
```

The plot of PC1 vs PC3 is as follows

```{r pca_13, echo=TRUE, eval=TRUE}
suppressWarnings(suppressMessages(iplot(pr$x[,1], pr$x[,3], as.numeric(as.factor(type_bodysite)),type_bodysite)))
```


Next, we pick the admixture/ Structure proportions from the above model fit for $K=6$ and then look at the t-SNE plot of these proportions.

```{r tsne_omega, echo=FALSE, eval=TRUE}
K=8
docweights <- as.matrix(read.table(paste0('../internal_data/hands_data/hands_topics_omega_',K,'.txt')));
#tsne_samples <- tsne(docweights,2);
#write.table(tsne_samples,paste0('../internal_data/hands_data/tsne_samples_',K,'.txt'));
tsne_samples <- as.matrix(read.table(paste0('../internal_data/hands_data/tsne_samples_',K,'.txt')));
options(warn=-1)
suppressWarnings(suppressMessages(iplot(tsne_samples[,1],tsne_samples[,2], as.numeric(as.factor(type_bodysite)),type_bodysite)))
```

The t-SNE plot does not seem to indicate the clusters that effectively, which we observed in Structure plot. We now consider the PCA plot to check if it shows any clustering between the fossa/crease or left hand/right hand. 

## Clustering of the full data 

We now consider the clustering of all the tissue samples together, not just Skin or crease/fossa. We check if the samples coming from other tissues lead to distinct clusters or not. 

```{r echo=TRUE, eval=FALSE}
full_gom <- CountClust::FitGoM(data, K=c(2,4,6,8), tol=0.1);
save(full_gom, file="../rdas/full_gom_fit.rda")
```


### K=6

```{r}
full_gom <- get(load("../rdas/full_gom_fit.rda"));
omega <- full_gom$clust_8$omega
theta <- full_gom$clust_8$theta

docweights_per_site_mean <- apply(omega, 2,
                                    function(x) { tapply(x, metadata$HMPbodysubsite, mean) })
ordering <- heatmap(docweights_per_site_mean)$rowInd

sites_ordered <- sort(unique(metadata$HMPbodysubsite))[ordering]


annotation <- data.frame(
    sample_id = paste0("X", 1:nrow(omega)),
    tissue_label = factor(metadata$HMPbodysubsite,
                          levels = rev(sites_ordered) )) 


StructureGGplot(omega = omega,
                annotation = annotation,
                palette = RColorBrewer::brewer.pal(8, "Accent"),
                yaxis_label = "hands body subsite",
                order_sample = TRUE,
                axis_tick = list(axis_ticks_length = .1,
                axis_ticks_lwd_y = .1,
                axis_ticks_lwd_x = .1,
                axis_label_size = 7,
                axis_label_face = "bold"))


```


### K=8

```{r}
full_gom <- get(load("../rdas/full_gom_fit.rda"));
omega <- full_gom$clust_8$omega
theta <- full_gom$clust_8$theta

docweights_per_site_mean <- apply(omega, 2,
                                    function(x) { tapply(x, metadata$HMPbodysubsite, mean) })
ordering <- heatmap(docweights_per_site_mean)$rowInd

sites_ordered <- sort(unique(metadata$HMPbodysubsite))[ordering]


annotation <- data.frame(
    sample_id = paste0("X", 1:nrow(omega)),
    tissue_label = factor(metadata$HMPbodysubsite,
                          levels = rev(sites_ordered) )) 


StructureGGplot(omega = omega,
                annotation = annotation,
                palette = RColorBrewer::brewer.pal(8, "Accent"),
                yaxis_label = "hands body subsite",
                order_sample = TRUE,
                axis_tick = list(axis_ticks_length = .1,
                axis_ticks_lwd_y = .1,
                axis_ticks_lwd_x = .1,
                axis_label_size = 7,
                axis_label_face = "bold"))


```



