---
title: "PCA of HMP data"
author: "Kushal K Dey"
date: "October 29, 2015"
output: 
html_document:
  toc: TRUE
---

## Objective 

We first perform PCA across all sites (and subsites) to see if the PCs separate out the different sites or not. Also intuitively samples coming from same individual should have similar OTU abundance patterns. So, if we zoom in further at each site and do PCA on the samples coming from that site, we should see samples from that site coming from same individual to be similar. 

## Loading the data 

```{r packages, message = FALSE, warning = FALSE}
library(metagenomeSeq)
library(qtlcharts)
```

```{r}
#setwd('/Users/kushal/Documents/metagenomics/project')
load(file = '../data/HMPvs35_nt100_MRexperiment.rdata')

hmp <- filterData(hmp, present=10)

counts <- t(MRcounts(hmp, norm=TRUE, log = FALSE))

metadata <- pData(hmp)
otu_data <- featureData(hmp)

dim(metadata)
#names(metadata)
#table(metadata$HMPbodysupersite)
```

## PCA of all sites (HMP)

### Supersite level

We assign to each point on the PCA plot, the supersite indicator from which it comes. 
```{r echo=TRUE, eval=TRUE}

cpm <- log(counts + 0.5)
pr.out=prcomp(cpm)
iplot(pr.out$x[,1], pr.out$x[,2], as.numeric(factor(metadata$HMPbodysupersite)), metadata$HMPbodysupersite)
```

### Subsite level

We assign to each point on the PCA plot, the subsite indicator from which it comes. 

```{r echo=TRUE, eval=TRUE}
iplot(pr.out$x[,1], pr.out$x[,2], as.numeric(factor(metadata$HMPbodysubsite)), metadata$HMPbodysubsite)

```

### Individual markers

We assign to each point on the PCA plot, the host id from which it comes (are the samples coming from same individual close??).

```{r echo=TRUE, eval=TRUE}
iplot(pr.out$x[,1], pr.out$x[,2], as.numeric(factor(metadata$hostsubjectid)), metadata$hostsubjectid)

```


## PC at each site separately

We next perform PCA on each of the sites and color by individuals to see if the individual effects are observed in some sites or not. If the PCs show the individual effect, then the topic model or relational topic model should also pick up that similarity. We check the individual effects for the $5$ supersites.

```{r echo=TRUE, eval=TRUE}

unique_supersites <- unique(metadata$HMPbodysupersite);

```

###  PCA on Oral

```{r echo=TRUE, eval=TRUE}
l <- 1
cpm_oral <- cpm[which(metadata$HMPbodysupersite == unique_supersites[l]),];
indiv_id <- metadata$hostsubjectid[which(metadata$HMPbodysupersite == unique_supersites[l])];
one_sample_indivs <- names(which(table(indiv_id)==1));
new_indiv_id <- indiv_id[- match(one_sample_indivs, indiv_id)]
cpm_oral <- cpm_oral[- match(one_sample_indivs, indiv_id),];
pr=prcomp(cpm_oral)
iplot(pr$x[,1], pr$x[,2], as.numeric(factor(new_indiv_id)), new_indiv_id)

```

### PCA on Skin

```{r echo=TRUE, eval=TRUE}
l <- 2
cpm_oral <- cpm[which(metadata$HMPbodysupersite == unique_supersites[l]),];
indiv_id <- metadata$hostsubjectid[which(metadata$HMPbodysupersite == unique_supersites[l])];
one_sample_indivs <- names(which(table(indiv_id)==1));
new_indiv_id <- indiv_id[- match(one_sample_indivs, indiv_id)]
cpm_oral <- cpm_oral[- match(one_sample_indivs, indiv_id),];
pr=prcomp(cpm_oral)
iplot(pr$x[,1], pr$x[,2], as.numeric(factor(new_indiv_id)), new_indiv_id)


```

### PCA on Urogenital Tract

```{r echo=TRUE, eval=TRUE}
l <- 3
cpm_oral <- cpm[which(metadata$HMPbodysupersite == unique_supersites[l]),];
indiv_id <- metadata$hostsubjectid[which(metadata$HMPbodysupersite == unique_supersites[l])];
one_sample_indivs <- names(which(table(indiv_id)==1));
new_indiv_id <- indiv_id[- match(one_sample_indivs, indiv_id)]
cpm_oral <- cpm_oral[- match(one_sample_indivs, indiv_id),];
pr=prcomp(cpm_oral)
iplot(pr$x[,1], pr$x[,2], as.numeric(factor(new_indiv_id)), new_indiv_id)


```

### PCA on Airways

```{r echo=TRUE, eval=TRUE}
l <- 4
cpm_oral <- cpm[which(metadata$HMPbodysupersite == unique_supersites[l]),];
indiv_id <- metadata$hostsubjectid[which(metadata$HMPbodysupersite == unique_supersites[l])];
one_sample_indivs <- names(which(table(indiv_id)==1));
new_indiv_id <- indiv_id[- match(one_sample_indivs, indiv_id)]
cpm_oral <- cpm_oral[- match(one_sample_indivs, indiv_id),];
pr=prcomp(cpm_oral)
iplot(pr$x[,1], pr$x[,2], as.numeric(factor(new_indiv_id)), new_indiv_id)



```

### PCA on Gastrointestinal Tract

```{r echo=TRUE, eval=TRUE}
l <- 5
cpm_oral <- cpm[which(metadata$HMPbodysupersite == unique_supersites[l]),];
indiv_id <- metadata$hostsubjectid[which(metadata$HMPbodysupersite == unique_supersites[l])];
one_sample_indivs <- names(which(table(indiv_id)==1));
new_indiv_id <- indiv_id[- match(one_sample_indivs, indiv_id)]
cpm_oral <- cpm_oral[- match(one_sample_indivs, indiv_id),];
pr=prcomp(cpm_oral)
iplot(pr$x[,1], pr$x[,2], as.numeric(factor(new_indiv_id)), new_indiv_id)


```

## Conclusion 

When PCA was done at the site level, the plot clearly shows that the samples coming from the same individual to cluster together in the PC space. However when PCA is applied on all sites together, it basically picks up the variation across the sites only.

```{r}
sessionInfo()
```

