---
title: "Explore relational topic model"
author: "Kushal K Dey, Joyce Hsiao, Joe Paulson"
date: "2015-09-24"
output: 
html_document:
  toc: TRUE
---

We use the relational topic model due to [Jonatha Huang](relational-topic) to uncover
underlying micriobiome clusters in the HMP data, in particular the crease subset. Specifically, we fit a topic model with the constraint that some samples come from
the sample individuals. The expectation is that the samples belong to the same
individuals are more likely to be grouped into the same clusters as opposed to 
samples that come from two different individuals. 

[relational-topic]: Chang, Jonathan and Blei, David M. Relational Topic Models for Document Networks. Artificial intelligence and statistics. 2009.

Note that in the literature, left hand generally is found to have very little in common
with right hand in the microbial community composition. However, I suspect that after
fitting a model with constraints, we will find more similarities between right and
left hand. Furthermore, do microbial communities share more similarities within 
and individual or between an individual??


## Set up

```{r load_data, message = FALSE, warning = FALSE}
library(metagenomeSeq)
library(maptpx)
#library(tsne)))
#library(qtlcharts)
#library(qvalue)
library(lda)
```


## Prepare data

```{r}
load(file = '../data/HMPvs35_nt100_MRexperiment.rdata')

hmp <- filterData(hmp, present=10)

counts <- t(MRcounts(hmp, norm=TRUE, log = FALSE))

metadata <- pData(hmp)
otu_data <- featureData(hmp)

counts_crease <- counts[which(metadata$HMPbodysubsite=="Right_Retroauricular_crease" |
                               metadata$HMPbodysubsite=="Left_Retroauricular_crease"), ]
metadata_crease <- metadata[which(metadata$HMPbodysubsite=="Right_Retroauricular_crease" |
                               metadata$HMPbodysubsite=="Left_Retroauricular_crease"), ]
table(metadata_crease$HMPbodysubsite)
```


Indiviudal labels

```{r}
person_label <- sapply(1:dim(metadata_crease)[1], function(x)  
                         strsplit( metadata_crease$Description[x],"_")[[1]][8] )

# Compare hostsubject id and also description 
which(person_label == "763435843")
cbind(person_label, metadata_crease$hostsubjectid)[c(1,156), ]

length(unique(metadata_crease$hostsubjectid))
length(unique(person_label))
```

## Relationtional topic model analysis

Prepare input

```{r}
## Choose only samples with 1 or 0 link in the data 
input_link <- lapply(1:length(metadata_crease$hostsubjectid), function(per_person) {
                          id <- metadata_crease$hostsubjectid[per_person]
                          exp <- metadata_crease$EXPERIMENTACCESSION[per_person]
                          pair <- which(metadata_crease$hostsubjectid == id)
                          pair[pair != per_person]
                          })
ii_subset <- which(sapply(input_link, function(xx) length(xx) < 2))

# counts entry needs to be a list 
counts_crease <- round(counts_crease)
counts_crease_subset <- counts_crease[ii_subset, ]
metadata_crease_subset <- metadata_crease[ii_subset, ]


input_crease_subset <- lapply(1:nrow(counts_crease_subset), function(per_person) {
                          rbind(as.integer( c(0: (ncol(counts_crease_subset) - 1 ) ) ),
                                as.integer(counts_crease_subset[per_person, ] ) ) 
                          })
input_link_subset <- lapply(1:length(metadata_crease_subset$hostsubjectid), function(per_person) {
                          id <- metadata_crease_subset$hostsubjectid[per_person]
                          exp <- metadata_crease_subset$EXPERIMENTACCESSION[per_person]
                          pair <- which(metadata_crease_subset$hostsubjectid == id)
                          as.integer(pair[pair != per_person] - 1)
                          })
input_genes <- colnames(counts_crease_subset)

stopifnot (max(unlist(input_link_subset)) == length(input_crease_subset) - 1)
stopifnot( min(unlist(input_link_subset)) == 0)
stopifnot( min(input_crease_subset[[1]] )  == 0)
```

Fit relational topic model

```{r}
if (!file.exists("../rdas/relation-topics-hmp.rda")) {
  rtm.model <- lapply(2:5, function(per_clust) {
                            rtm.collapsed.gibbs.sampler(input_crease_subset,
                                                        input_link_subset,
                                                        per_clust,
                                                        vocab = input_genes,
                                                        num.iterations = 35,
                                                        alpha = 0.1, 
                                                        eta = 0.1, beta = 3)
                        })
save(rtm.model, file = "../rdas/relation-topic-hmp.rda")
} else {
  load("../rdas/relation-topic-hmp.rda")  
}
```

```{r}
# document_sums: each entry is a numeric proportional
# to the probability of seeing a topic (row) conditioned on the document column

topic_probs <- lapply(1:length(rtm.model), function(per_clust) {
                      t(t(rtm.model[[per_clust]]$document_sums)/
                            colSums(rtm.model[[per_clust]]$document_sums))
                      })
dim(topic_probs[[1]])
```

Topic results

* When assume 2 underlying clusters

```{r}
# Find the maximum topic probability for each document
ii_topic <- apply(rtm.model[[1]]$document_sums, 2, which.max)
unique_subject <- table(metadata_crease_subset$hostsubjectid) 
unique_subject <- names(unique_subject)[unique_subject==2]
subjects <- metadata_crease_subset$hostsubjectid

topic_match_sort <- lapply(unique_subject, function(xx) {
    ii_subject <- which(subjects %in% xx)
    ii_topic[ii_subject]
})
topic_match_sort

table(sapply(topic_match_sort, function(xx) xx[1] == xx[2]))
```

Most individuals have similar micriobial community on the left versus
on the right hand. 



* When assume 5 underlying clusters

```{r}
# Find the maximum topic probability for each document
ii_topic <- apply(rtm.model[[4]]$document_sums, 2, which.max)
unique_subject <- table(metadata_crease_subset$hostsubjectid) 
unique_subject <- names(unique_subject)[unique_subject==2]
subjects <- metadata_crease_subset$hostsubjectid

topic_match_sort <- lapply(unique_subject, function(xx) {
    ii_subject <- which(subjects %in% xx)
    ii_topic[ii_subject]
})
topic_match_sort

table(sapply(topic_match_sort, function(xx) xx[1] == xx[2]))
```



Structure plot

```{r}
nclust <- 2:5
ii_order <- order(metadata_crease_subset$hostsubjectid)
for (per_clust in 1:1) {
      clust <- nclust[per_clust]
      docweights <- topic_probs[[per_clust]]
      #ii_order <- order(docweights[1, ], docweights[2, ])
      docweights_ordered <- docweights[ , ii_order]
      barplot( docweights_ordered, 
               col = 2:(clust+1), axisnames = F,
               space = 0, border=NA,
               main = paste("No. of clusters=", clust), las = 1,
               ylim = c(0,1), cex.axis = 0.3, cex.main=1.4)
#      labels <- match(unique(type_ordered[,1]), type_ordered[,1])
#       abline(v=labels-1, lwd =3)
#       labels_low <- labels-1
#       labels_up <- c(labels_low[2:length(labels_low)], dim(docweights_ordered)[1])
#       mid_point <- labels_low +0.5*(labels_up-labels_low)
#       axis(1, at = mid_point, unique(type_ordered[,1]), las = 2, cex.axis=0.3)
}
```



## Session information

```{r}
sessionInfo()
```